Prognostic and biological relevance of collagen-related genes in prostate cancer

### 2025-02-24 paper/markdown/manuscript_figures_tables.Rmd
Supplementary Methods
Supplementary Methods and a backbone for the main methods. Changes in labels and legends of Supplementary Figures

>>>>

# Methods

Analyses of collagen-related transcriptome and survival in transcriptomic cohorts were done with R version 4.2.3 (R foundation). 
Their details are provided in __Supplementary Methods__.

## Transcriptome data sets

Seven publicly available transcriptome data sets of primary prostate cancer samples, 
the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] (n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, RNA sequencing), 
GSE16560 [@Sboner2010] (n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray)
GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), and 
DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
were re-analyzed. 
Selection criteria for the transcriptomic cohorts were: information of biochemical relapse and relapse-free survival or death and overall survival, and availability of expression data for `r nrow(globals$genes_interest)` collagen-related transcript investigated in our previous collagen project [@Heidegger2024].

Normalized gene expression data and clinical information were fetched from the cBioportal repository (DKFZ, TCGA) or Gene Expression Omnibus (GEO). 
The microarray signals of RNA sequencing counts were analyzed as $log_2$-transformed counts adjusted for differences in distribution between the cohorts with ComBat algorithm [@Leek2012]. 
ComBat-adjusted expression and clinical information for the `r globals$study_labels[geo_pool$studies] %>% collapse_and` studies was merged into a 'pooled GEO cohort'. 
In machine learning models of BCR-free survival, the pooled GEO cohort was used for training, and TCGA and DKFZ served as validation cohorts. 
The GSE16560 cohort split into a training and test subset in a 2:1 ratio was used for modeling of overall survival. 
Characteristic of the cohorts and a list of the collagen-related genes of  interest are are provided in __Supplementary Tables S1 - S4__.

## Co-expression network analyses

## Univariable modeling of BCR-free survival

## Multi-parameter modeling of BCR-free and overall survival

## Data and code availability

Publicly available data sets were analyzed. 
Formatted data sets used for analyses will be made available upon request to the corresponding author. 
The R analysis pipeline for the transcriptome data is available from GitHub (https://github.com/PiotrTymoszuk/BCR_collagen).
